---
title: "Death Rates for Suicide, by Sex, Race, Hispanic Origin, and Age"
format: html
editor: visual
author: "Hely Bhavsar, Sheridan Harney, Jasjeevan Kaur"
---

<https://catalog.data.gov/dataset/death-rates-for-suicide-by-sex-race-hispanic-origin-and-age-united-states-020c1>

## Introduction

This is a project analyzing suicide death rates.

## Death Rates Questions

1. Has the suicide estimate changed over time?
2. How do suicide rate estimates differ across young adult age groups?


## Running Code

```{r}
library(readr)

Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States <- read_csv("data/Death_rates_for_suicide__by_sex__race__Hispanic_origin__and_age__United_States.csv")

install.packages("ggplot2")

library(ggplot2)
```

```{r}
ggplot(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States, aes(x = ESTIMATE)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") +
  labs(title = "Distribution of Suicide Rates", x = "Suicide Rate (per 100,000)", y = "Frequency")

  ggplot(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States, aes(x = STUB_LABEL, y = INDICATOR, fill = STUB_LABEL)) +
    geom_boxplot() +
    labs(title = "Suicide Rates by Race", x = "race", y = "Suicide Rate (per 100,000)")
  
```

#Mean, Median, Mode, Standard Deviation, Min and Max
```{r}
install.packages("dplyr")
  library(dplyr)
  
  # Mean
mean(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States$ESTIMATE, na.rm = TRUE)

# Median
median(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States$ESTIMATE, na.rm = TRUE)

# Mode (base R way)
get_mode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}
get_mode(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States$ESTIMATE)

# Standard Deviation
sd(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States$ESTIMATE, na.rm = TRUE)

# Minimum and Maximum
min(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States$ESTIMATE, na.rm = TRUE)
max(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States$ESTIMATE, na.rm = TRUE)

    colnames(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States)
```

##Anova
```{r}
str(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States)         
    unique(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States$STUB_LABEL) 
  
  
    aov_result <- aov(ESTIMATE ~ STUB_LABEL, data = Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States)
    summary(aov_result)
```


```{r}

    correlation_result <- cor.test(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States$"AGE_NUM", Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States$ESTIMATE)
    print(correlation_result)
    
    
    Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States$STUB_LABEL <- as.factor(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States$STUB_LABEL)
    
  
    model <- lm(ESTIMATE ~ STUB_LABEL, data = Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States)

    summary(model)
    
    colnames(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States)
```

##Questions

1.  Has the suicide estimate changed over time?

```{r}

    library(ggplot2)
library(dplyr)  

# First: create male_female_data
male_female_data <- Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States %>%
  filter(STUB_LABEL %in% c("Male", "Female"))


lm_model <- lm(ESTIMATE ~ YEAR, data = male_female_data)
summary(lm_model)


ggplot(male_female_data, aes(x = YEAR, y = ESTIMATE)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Trend in Suicide Estimates Over Time", x = "Year", y = "Estimate") +
  theme_minimal()

```

Interpretation:

Linear regression assesses whether suicide estimates are increasing or decreasing over time. The slope coefficient of YEAR tells us the direction of the trend, and the p-value indicates if the trend is significant.

Result :

As the years increase, the estimated number of suicides slightly decreases on average. However, the data points are very spread out, meaning that year is not a strong predictor of suicide estimates by itself. The relationship is weak, even though it trends downward.

2.  How do suicide rate estimates differ across young adult age groups?

```{r}

unique(Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States$AGE)

age_data <- Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States

library(dplyr)

age_data <- filter(
  Death_rates_for_suicide_by_sex_race_Hispanic_origin_and_age_United_States,
  AGE %in% c("15-24 years", "25-34 years", "35-44 years")
)

table(age_data$AGE)


age_data$AGE <- as.factor(age_data$AGE)


anova_age <- aov(ESTIMATE ~ AGE, data = age_data)
summary(anova_age)


library(ggplot2)
ggplot(age_data, aes(x = AGE, y = ESTIMATE, fill = AGE)) +
  geom_boxplot() +
  labs(title = "Suicide Estimates by Age Group", x = "Age Group", y = "Estimate") +
  theme_minimal()





```

Interpretation:

The ANOVA tested whether there were significant differences in average suicide rate estimates among three young adult age groups: 15–24, 25–34, and 35–44 years. The resulting p-value from the ANOVA summary indicates \[whether or not\] these differences are statistically significant.

Result:

Based on the box plot, age group 35-44 years tends to have higher estimates than 15–24, 25–34 year age group.

Conclusion:

In order to begin our analysis, we first constructed a histogram to examine the distribution of suicide estimates. The histogram indicated that the data was left-skewed, i.e., while most of the suicide estimates were quite low, there were certain groups who had much higher values. We then created boxplots in order to examine the spread of suicide estimates across groups. The boxplots showed clear differences between age groups, with older men—especially those 35 years and older—having greater suicide estimates compared to younger age groups. The visualizations helped us to know which groups were most vulnerable.
Once we graphed the data, we calculated the mean, median, mode, standard deviation, minimum, and maximum to provide measures of central tendency and variability for the suicide estimates. The mean was substantially greater than the median, again reflecting the right-skewed distribution that we had observed from the histogram. To statistically determine whether there were any differences among suicide estimates within demographic groups, we conducted an ANOVA (Analysis of Variance). The ANOVA test indicated a statistically significant difference, suggesting that suicide estimates varied significantly between groups. The results, combined with the visualizations, suggest that age, gender, and race/ethnicity influence suicide estimates, underscoring the need for targeted mental health interventions among high-risk groups.
